---
title: 'The effect of fast starting on finish times: Boston Marathon data analysis'
author: "Ana Belen Barcenas J."
date: "11/28/2018"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
setwd("~/Documents/MIDS/Modeling and Repr of Data/Final project/Boston_marathon")
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results="hide"}
# install.packages("lubridate")
library(lubridate)
library(ggplot2)
library(RColorBrewer)
library(dplyr)
library(splines)
library(scales)
library(Metrics)
library(sjstats)
library(mgcv)
library(caret)
library(tidyverse)
library(fiftystater)
```


Importing the datasets from Boston Marathon results in 2015, 2016, and 2017:
```{r}
BM_2017=read.csv("~/Documents/MIDS/Modeling and Repr of Data/Final project/Boston_marathon/dataset/marathon_results_2017.csv")
BM_2016=read.csv("~/Documents/MIDS/Modeling and Repr of Data/Final project/Boston_marathon/dataset/marathon_results_2016.csv")
BM_2015=read.csv("~/Documents/MIDS/Modeling and Repr of Data/Final project/Boston_marathon/dataset/marathon_results_2015.csv")
```

Merge the datasets from 2015, 2016, 2017. But first, let's add a column of the specific year and weather during the marathon.
```{r}
BM_2017$X <- NULL
BM_2017$X.1 <- NULL
BM_2017$year <- 2017
BM_2017$temp_f <- 73

BM_2016$X <- NULL
BM_2016$year <- 2016
BM_2016$temp_f <- 61

BM_2015$X <- NULL
BM_2015$X.1 <- NULL
BM_2015$year <- 2015
BM_2015$temp_f <- 46

boston_m <- rbind(BM_2017, BM_2016, BM_2015)
boston_m$Proj.Time <- NULL
```

There are some missing values but the proportion is less than 1%. I'll get ride of those observations.
```{r}
boston_m[boston_m=='-']=NA
boston_m <- boston_m[complete.cases(boston_m),]
```


Let's transform some variables classes to what we'll need for the model
```{r, results = 'hide'}
boston_m$OfficialTsec2 <- as.POSIXct(boston_m$Official.Time,format='%H:%M:%S')
substr(boston_m$OfficialTsec2,12,19)

# Race times per cohort - convert them in seconds
boston_m$X5K <- hms(as.character(boston_m$X5K))
boston_m$X5Ksec <- hour(boston_m$X5K)*60*60 + minute(boston_m$X5K)*60 + second(boston_m$X5K)

boston_m$X10K <- hms(as.character(boston_m$X10K))
boston_m$X10Ksec <- hour(boston_m$X10K)*60*60 + minute(boston_m$X10K)*60 + second(boston_m$X10K)

boston_m$X15K <- hms(as.character(boston_m$X15K))
boston_m$X15Ksec <- hour(boston_m$X15K)*60*60 + minute(boston_m$X15K)*60 + second(boston_m$X15K)

boston_m$X20K <- hms(as.character(boston_m$X20K))
boston_m$X20Ksec <- hour(boston_m$X20K)*60*60 + minute(boston_m$X20K)*60 + second(boston_m$X20K)

boston_m$Half <- hms(as.character(boston_m$Half))
boston_m$Halfsec <- hour(boston_m$Half)*60*60 + minute(boston_m$Half)*60 + second(boston_m$Half)

boston_m$X25K <- hms(as.character(boston_m$X25K))
boston_m$X25Ksec <- hour(boston_m$X25K)*60*60 + minute(boston_m$X25K)*60 + second(boston_m$X25K)

boston_m$X30K <- hms(as.character(boston_m$X30K))
boston_m$X30Ksec <- hour(boston_m$X30K)*60*60 + minute(boston_m$X30K)*60 + second(boston_m$X30K)

boston_m$X35K <- hms(as.character(boston_m$X35K))
boston_m$X35Ksec <- hour(boston_m$X35K)*60*60 + minute(boston_m$X35K)*60 + second(boston_m$X35K)

boston_m$X40K <- hms(as.character(boston_m$X40K))
boston_m$X40Ksec <- hour(boston_m$X40K)*60*60 + minute(boston_m$X40K)*60 + second(boston_m$X40K)

boston_m$Official.Time <- hms(as.character(boston_m$Official.Time))
boston_m$OfficialTsec <- hour(boston_m$Official.Time)*60*60 + minute(boston_m$Official.Time)*60 + second(boston_m$Official.Time)

# Pace per cohort - convert them in seconds
boston_m$X5Kpace <- boston_m$X5Ksec/5
boston_m$X10Kpace <- (boston_m$X10Ksec-boston_m$X5Ksec)/5
boston_m$X15Kpace <- (boston_m$X15Ksec-boston_m$X10Ksec)/5
boston_m$X20Kpace <- (boston_m$X10Ksec-boston_m$X15Ksec)/5
boston_m$Halfpace <- (boston_m$Halfsec-boston_m$X20Ksec)/(21.1-20)
boston_m$X25Kpace <- (boston_m$X25Ksec-boston_m$Halfsec)/(25-21.1)
boston_m$X30Kpace <- (boston_m$X30Ksec-boston_m$X25Ksec)/5
boston_m$X35Kpace <- (boston_m$X35Ksec-boston_m$X30Ksec)/5
boston_m$X40Kpace <- (boston_m$X40Ksec-boston_m$X35Ksec)/5
boston_m$X42.2Kpace <- (boston_m$OfficialTsec-boston_m$X40Ksec)/(42.2-40)
boston_m$mean_pace <- boston_m$OfficialTsec/42.2
```

Create a dummy variable for each gender
```{r}
# Gender
boston_m$M.F <- as.character(boston_m$M.F)
boston_m$Female[boston_m$M.F =='F'] <- 1
boston_m$Female[boston_m$M.F =='M'] <- 0
```

Create a variable of the type of runner: elite, recreational, walker.
```{r}
boston_m <- boston_m %>%
  mutate(category = ifelse((Female==1 & OfficialTsec<=11100), "elite",
         ifelse((Female==0 & OfficialTsec<=9000), "elite", "recreational")))
```

Percentage of runners in each category
```{r}
proportion_cat <- round((table(boston_m$category) / nrow(boston_m)),2)*100
proportion_cat
```
Percentage of gender
```{r}
proportion_gen <- round((table(boston_m$M.F) / nrow(boston_m)),2)*100
proportion_gen
```
```{r}
mean(boston_m$OfficialTsec)/60
```

Create the relative pace variable of the first 5 kms: 100*((mean_pace - X5Kpace)/mean_pace) and last kms
```{r}
boston_m <- boston_m %>%
  mutate(RPace5K = ((boston_m$mean_pace - boston_m$X5Kpace)/boston_m$mean_pace)*100)

boston_m <- boston_m %>%
  mutate(RPace42.2K = ((boston_m$mean_pace - boston_m$X42.2Kpace)/boston_m$mean_pace)*100)
```



# EXPLORATORY DATA ANALYSIS


1. Country of origin, count vs finish time / USA state of origin, count vs finish time


# Let's count how many runners do we have per state
```{r}
boston_m$id <- rownames(boston_m)

require(data.table)
setDT(boston_m)[, runners := uniqueN(id), by = State]
user_state <- boston_m %>%
  dplyr::select(State, runners)
user_state <- unique(user_state)
```

# USA states coordinates
```{r}
mapdata <- read.csv("~/Documents/MIDS/Modeling and Repr of Data/Final project/Boston_marathon/dataset/usa_states.csv", header=TRUE, stringsAsFactors=FALSE) 
colnames(mapdata)[1] <- "State"
```

# Let's calculate the average click apply and the avg finish time in each state
```{r}
setDT(boston_m)[, avg_finish_time := .(mean(OfficialTsec)), by = State]
```

```{r}
runners_state <- boston_m %>%
  dplyr::select(State, avg_finish_time)
runners_state <- unique(runners_state)
```

# Let's join users count, state and coordinates, and click apply avg per person
```{r}
x2 <- inner_join(runners_state, mapdata, by = "State")
x3 <- inner_join(x2,user_state, by="State") # Merge total click applies by state
# we get ride of states whose coordinates are outside the USA map
x3 <- x3 %>%
  dplyr::select(State,runners,avg_finish_time,location,lat,lon)
x3 <- unique(x3)
```
# Let's plot the map!!
```{r}
data("fifty_states")
ggplot() + geom_polygon( data=fifty_states, aes(x=long, y=lat, group = group),color="white", fill="grey10" )

ggplot() + geom_polygon(data=fifty_states, aes(x=long, y=lat, group = group),color="white", fill="grey92" ) +
  geom_point(data=x3, aes(x=lon, y=lat, size = runners, color=avg_finish_time/60)) +
  scale_size(name="", range = c(2, 20)) +
  guides(size=guide_legend("Runners")) +
  labs(col="Average finish time (mins.)") +
  theme_void()
```




2. Gender vs. finish time

```{r}
# Define color scale
myColors_g <- c("#66CC99","#006699")
names(myColors_g) <- levels(boston_m$M.F)

# Density plots of running times to define elite, recreational, and walkers runners
g <- ggplot(boston_m, aes(x=OfficialTsec2, colour = M.F, fill=M.F)) + geom_density(alpha=.7)
g1 <- g + scale_colour_manual(name = "Gender",values = myColors_g) + 
  scale_fill_manual(name = "Gender",values = myColors_g) + labs(x = "Official Marathon Time") +
  theme_minimal() +
  scale_x_datetime(breaks = pretty_breaks(9))
g1
```

3.1. Age vs finish time (scatterplot or line) : Recreational

```{r}

require(data.table)
setDT(boston_m)[, mean_finish_time_age := .(mean(OfficialTsec)), by = Age]

myColors_g <- c("#66CC99","#006699")
names(myColors_g) <- levels(boston_m$M.F)

ggplot(boston_m[category=="recreational",], aes(x=Age, y=mean_finish_time_age, color=M.F)) +
  geom_point(shape=1) + geom_smooth() + theme_minimal() + theme(legend.position="bottom") +
  scale_colour_manual(name = "Gender",values = myColors_g) + 
  scale_fill_manual(name = "Gender",values = myColors_g) + 
  labs(x = "Age", y = "Mean finish time (seconds)")
```

3.2. Age vs finish time (scatterplot or line) : Elite
```{r}
ggplot(boston_m[category=="elite",], aes(x=Age, y=mean_finish_time_age, color=M.F)) +
  geom_point(shape=1) + geom_smooth() + theme_minimal() +  theme(legend.position="bottom") +
  scale_colour_manual(name = "Gender",values = myColors_g) + 
  scale_fill_manual(name = "Gender",values = myColors_g) + 
  labs(x = "Age", y = "Mean finish time (seconds)")
```


4. Start pace vs. finish time

```{r}
plot(y = boston_m$OfficialTsec, x = boston_m$X5Kpace, xlab = "X5Kpace", ylab = "OfficialTsec")
```

#elite
```{r}
plot(y = boston_m[boston_m$category=="elite",]$OfficialTsec, x = boston_m[boston_m$category=="elite",]$X5Kpace, xlab = "X5Kpace", ylab = "OfficialTsec")
```


5. Relative start pace vs. finish time
```{r}
plot(y = boston_m[1:5000,]$OfficialTsec, x = boston_m[1:5000,]$RPace5K, xlab = "RPace5K", ylab = "OfficialTsec")
```

6. Relative start pace vs. average finish time

The relationship does not look clear, I'll take the average finish time every 1% difference in relative start (5kms) pace.
```{r}
require(data.table)
boston_m <- boston_m %>%
  mutate(round_RPace5K=round(RPace5K,1))
setDT(boston_m)[, mean_finish_time_5pace := .(mean(OfficialTsec)), by = round_RPace5K]
```

6.1. Recreational runners by gender

Let's see the relationship between finish time and relative pace at the beginning of the race (5km)
```{r}
myColors_g <- c("#66CC99","#006699")
names(myColors_g) <- levels(boston_m$M.F)

ggplot(boston_m[category=="recreational",], aes(x=RPace5K, y=mean_finish_time_5pace, color=M.F)) +
  geom_point(shape=1) + geom_smooth() + theme(legend.position="top") + theme_minimal() +
  scale_colour_manual(name = "Gender",values = myColors_g) + 
  scale_fill_manual(name = "Gender",values = myColors_g) + 
  labs(x = "Relative Pace - 5km", y = "Mean finish time (seconds)")
```

6.2. Elite runners by gender
```{r}
ggplot(boston_m[category=="elite",], aes(x=RPace5K, y=mean_finish_time_5pace, color=M.F)) +
  geom_point(shape=1) + geom_smooth() + theme(legend.position="top") + theme_minimal() +
  scale_colour_manual(name = "Gender",values = myColors_g) + 
  scale_fill_manual(name = "Gender",values = myColors_g) + 
  labs(x = "Relative Pace - 5km", y = "Mean finish time (seconds)")
```


7.1. Recreational runners by gender an start pace (draw line in the 0 and slow/faster starters)

Let's see the distribution of runners vs relative pace for recreational runners
```{r}
d <- ggplot(boston_m[boston_m$category=="recreational",], 
            aes(x=RPace5K, colour = M.F, colour=M.F)) + geom_density()
d + scale_colour_manual(name = "Gender",values = myColors_g) + 
  scale_fill_manual(name = "Gender",values = myColors_g) + labs(x = "Relative Pace - 5km") +
  theme_minimal()
```

7.2. Elite runners by gender

Let's see the distribution of runners vs relative pace for ELITE runners
```{r}
d <- ggplot(boston_m[boston_m$category=="elite",], 
            aes(x=RPace5K, colour = M.F, colour=M.F)) + geom_density()
d + scale_colour_manual(name = "Gender",values = myColors_g) + 
  scale_fill_manual(name = "Gender",values = myColors_g) + labs(x = "Relative Pace - 5km") +
  theme_minimal()
```

8. Relative start pace vs. relative finish pace by gender. Are women too cautiously?
```{r}

setDT(boston_m)[, mean_finish_pace := .(mean(RPace42.2K)), by = round_RPace5K]


myColors_g <- c("#66CC99","#006699")
names(myColors_g) <- levels(boston_m$M.F)

ggplot(boston_m[category=="recreational",], aes(x=RPace5K, y=mean_finish_pace, color=M.F)) +
  geom_point(shape=1) + geom_smooth() + theme(legend.position="top") + theme_minimal() +
  scale_colour_manual(name = "Gender",values = myColors_g) + 
  scale_fill_manual(name = "Gender",values = myColors_g) + 
  labs(x = "Relative start pace - 5km", y = "Relative finish pace - 5km")
```


9. Relative finish pace vs. avg finish time?
I'll take the average finish time every 1% difference in relative finish pace.
```{r}
require(data.table)
boston_m <- boston_m %>%
  mutate(round_RPace10K=round(RPace42.2K,1))
setDT(boston_m)[, mean_finish_time_10pace := .(mean(OfficialTsec)), by = round_RPace10K]


myColors_g <- c("#66CC99","#006699")
names(myColors_g) <- levels(boston_m$M.F)

ggplot(boston_m[category=="recreational",], aes(x=RPace42.2K, y=mean_finish_time_10pace, color=M.F)) +
  geom_point(shape=1) + geom_smooth() + theme(legend.position="top") + theme_minimal() +
  scale_colour_manual(name = "Gender",values = myColors_g) + 
  scale_fill_manual(name = "Gender",values = myColors_g) + 
  labs(x = "Relative Pace - 42.2km", y = "Mean finish time (seconds)")
```


# MODEL

# Split the data into training and test set
```{r}
set.seed(123)
training.samples <- boston_m$OfficialTsec %>%
  createDataPartition(p = 0.8, list = FALSE) 
train.data  <- boston_m[training.samples, ]
test.data <- boston_m[-training.samples, ]
```

1. Linear regression: finish pace vs. start pace

```{r}
reg1 <- lm(RPace42.2K ~ as.factor(year) + Age + Female + RPace5K,
           data = train.data[train.data$category=="recreational",])
summary(reg1)
```

2. Linear regression: finish time vs. start pace

```{r}
reg2 <- lm(OfficialTsec ~ as.factor(year) + Age + Female + RPace5K,
           data = train.data[train.data$category=="recreational",])
summary(reg2)
```

3.1. Non-linear model - Cubic splines: finish time vs. start pace RECREATIONAL

```{r}
reg3 <- lm(OfficialTsec ~ as.factor(year) + Age + Female + bs(RPace5K, degree = 2, knots = 2),
           data = train.data[train.data$category=="recreational",])
summary(reg3)
```

Conficence intervals:
```{r}
confint(reg3, level=0.95)
```

How well the splines fits the data?
```{r}
ggplot(train.data[train.data$category=="recreational",], aes(RPace5K, OfficialTsec) ) +
  geom_point() +
  stat_smooth(method = lm, formula = y ~ splines::bs(x, degree = 2, knots = 3))

```

3.1.1 Model diagnostics

```{r}
par(mfrow=c(2,2))
plot(reg3)
par(mfrow=c(1,1))
```

```{r, fig.width=7, fig.height=4}
par(mfrow=c(2,2)) 

boxplot(reg3$residuals~train.data[train.data$category=="recreational",]$year, ylab = "Residuals", xlab = "Year")
boxplot(reg3$resid~train.data[train.data$category=="recreational",]$Female, ylab = "Residuals", xlab = "Gender")

plot(reg3$resid, x=train.data[train.data$category=="recreational",]$Age, ylab = "Residuals", xlab = "Age")
abline(0,0)
plot(reg3$resid, x=train.data[train.data$category=="recreational",]$RPace5K, ylab = "Residuals", xlab = "Pace")
abline(0,0)
```

3.1.2. Outliers, leverage, and/or influential points.

Now that I have decided about what predictors, interactions, and transformations I will include in the model, I'll double check if there is any outlier, leverage points, and/or influential points to pay attention in.

```{r, fig.width=4, fig.height=2, fig.align="center"}
library(MASS)
leverage = hatvalues(reg3)
cooks = cooks.distance(reg3)
new_dataset = cbind(train.data, leverage, cooks)

hist(leverage, main = "Leverage values for Boston Marathon regression")
```
```{r, fig.width=4, fig.height=2, fig.align="center"}
hist(cooks, main = "Cook's distances for Boston Marathon regression")
```

3.1.3 Model predictions and performance
```{r}
predictions <- reg3 %>% predict(test.data)
# Model performance
data.frame(
  RMSE = RMSE(predictions, test.data$OfficialTsec),
  R2 = R2(predictions, test.data$OfficialTsec)
)
```




3.2. Non-linear model - Cubic splines: finish time vs. start pace ELITE

```{r}
reg4 <- lm(OfficialTsec ~ as.factor(year) + Age + Female + bs(RPace5K, degree = 2, knots = 0),
           data = train.data[train.data$category=="elite",])
summary(reg4)
```

Conficence intervals:
```{r}
confint(reg3, level=0.95)
```

How well the splines fits the data? 
```{r}
ggplot(train.data[train.data$category=="elite",], aes(RPace5K, OfficialTsec) ) +
  geom_point() +
  stat_smooth(method = lm, formula = y ~ splines::bs(x, degree = 2, knots = 0))

```


3.2.1 Model diagnostics
```{r, fig.width=7, fig.height=4}
par(mfrow=c(2,2)) 

boxplot(reg4$residuals~train.data[train.data$category=="elite",]$year, ylab = "Residuals", xlab = "Year")
boxplot(reg4$resid~train.data[train.data$category=="elite",]$Female, ylab = "Residuals", xlab = "Gender")

plot(reg4$resid, x=train.data[train.data$category=="elite",]$Age, ylab = "Residuals", xlab = "Age")
abline(0,0)
plot(reg4$resid, x=train.data[train.data$category=="elite",]$RPace5K, ylab = "Residuals", xlab = "Pace")
abline(0,0)
```

3.2.2. Outliers, leverage, and/or influential points.

Now that I have decided about what predictors, interactions, and transformations I will include in the model, I'll double check if there is any outlier, leverage points, and/or influential points to pay attention in.

```{r, fig.width=4, fig.height=2, fig.align="center"}
leverage = hatvalues(reg4)
cooks = cooks.distance(reg4)
new_dataset = cbind(train.data, leverage, cooks)

hist(leverage, main = "Leverage values for Boston Marathon regression")
```
```{r, fig.width=4, fig.height=2, fig.align="center"}
hist(cooks, main = "Cook's distances for smoking regression")
```


3.2.3 Model predictions and performance
```{r}
predictions2 <- reg4 %>% predict(test.data)
# Model performance
data.frame(
  RMSE = RMSE(predictions2, test.data$OfficialTsec),
  R2 = R2(predictions2, test.data$OfficialTsec)
)
```





